Detailed disease progression of 213 patients hospitalized with Covid-19 in the Czech Republic: An exploratory analysis

### 2020-11-26 manuscript/manuscript.Rmd
Cleaning up, finishing touches on manuscript, work on supplement

---
title: "Detailed disease progression of 213 patients hospitalized with Covid-19 in the Czech Republic: An exploratory analysis" 
output: 
  #rticles::arxiv_article
  #pdf_document
  # bookdown::pdf_document2:
  #    toc: false
  bookdown::word_document2: 
    toc: false
    keep_md: true
    reference_docx: "output_template.docx"
abstract: "We collected a multi-centric retrospective dataset of patients who were admitted to N hospitals in Czech Republic and tested positive for SARS-CoV-2. The dataset contains baseline patient characteristics, breathing support required, pharmacological treatment received and multiple markers on daily resolution. Patients in the dataset were treated with Hydroxychloroquine (N = 108), Azithromycin (N = 72), Favipiravir (N = 9), convalescent plasma (N = 7), Dexamethasone (N = 4) and Remdesivir (N = 3), often in combination. To explore association between treatments and patient outcomes we performed multiverse analysis, observing how the conclusions change between defensible choices of statistical model, predictors included in the model and other analytical degrees of freedom.  Weak evidence to constrain the potential efficacy of Azithromycin and Favipiravir can be extracted from the data. Additionally we performed external validation of several proposed prognostic models for Covid-19 severity showing that they mostly perform unsatisfactorily on our dataset."
bibliography: Covid19.json
date: "Version: `r format(Sys.time(), '%Y-%m-%d')`"
author: "`r { source(here::here('manuscript', 'authors.R')); get_authors_string() }`"
---

@@

The Covid-19 pandemic, caused by severe acute respiratory syndrome coronavirus (SARS-CoV-2) has, as of November 2020 led to over 46 million cases and over 1.1 million deaths. Proposed treatments include antivirals approved for other indications (Chloroquine, Hydroxychloroquine, Lopinavir/Ritonavir, Remdesivir, Favipiravir, Umifenovir), Azithromycin, corticosteroids, immunoglobulins, Tocilizumab and convalescent plasma [@http://zotero.org/users/5567156/items/8XWTUMUI; @http://zotero.org/users/5567156/items/6Q8Z8VJR]. Currently, the only treatment that exhibited positive effects in randomized trials is the corticosteroid Dexamethasone for severe cases [@http://zotero.org/users/5567156/items/IW6VZ53Q; @http://zotero.org/users/5567156/items/5VAKEW58], while results on Remdesivir have been conflicting [@http://zotero.org/users/5567156/items/MFS3IPPU; @http://zotero.org/users/5567156/items/NSTX4EZM]. Despite initial promise, Hydroxychloroquine has failed to show benefit for both hospitalized patients [@http://zotero.org/users/5567156/items/9X3UH8ZM; @http://zotero.org/users/5567156/items/NSTX4EZM], as early treatment [@http://zotero.org/users/5567156/items/G6KZKVMI], and as both pre- and post-exposure prophylaxis [@http://zotero.org/users/5567156/items/7C23YCV5; @http://zotero.org/users/5567156/items/GMR9FX9Z; @http://zotero.org/users/5567156/items/H4CLJNUR]. Further, no improvement could be discerned when adding Azithromycin to Hydroxychloroquine therapy [@http://zotero.org/users/5567156/items/BQLRZGTQ]. Similarly, Lopinavir/Ritonavir also did not show benefit [@http://zotero.org/users/5567156/items/6I6GAVWU]. No benefit was observed for convalescent plasma [@http://zotero.org/users/5567156/items/UEXLDJCB; @http://zotero.org/users/5567156/items/BHGT88US].

Interestingly, in observational studies, Hydroxychloroquine was often found to be associated with better outcomes [e. g., @http://zotero.org/users/5567156/items/WZPNBRYN; @http://zotero.org/users/5567156/items/4HR7MUJ6; @http://zotero.org/users/5567156/items/NXC9CDI9].

Favipiravir is one of the less explored experimental treatments, with only one larger study known to us [@http://zotero.org/users/5567156/items/ISLSL7UC], the study however has many potentially severe problems [@http://zotero.org/users/5567156/items/7HDXLWEA]. Even considering studies or case series with only several treated patients, very few are available [@http://zotero.org/users/5567156/items/F3VNTQ92; @http://zotero.org/users/5567156/items/FSUQMUP8; @http://zotero.org/users/5567156/items/894H9N9B]. None of the results allow strong conclusions about benefits or harms of Favipiravir. For additional summary of treatments under investigation see [@http://zotero.org/users/5567156/items/YP62P6XJ].

@@

# Methods

@@

n_ecmo <- data$marker_data_wide %>% filter(breathing == "ECMO") %>% pull(patient_id) %>% unique() %>% length()
n_nippv <- data$marker_data_wide %>% filter(breathing == "NIPPV") %>% pull(patient_id) %>% unique() %>% length()
```

A convenience sample of patients from `r n_hospitals` sites was collected. For each site, the dataset contains all patients hospitalized in the participating wards over the data collection period. The data collection started at the onset of the Covid-19 pandemic, but the end data for collection differs between sites.   Patients over the age of 18 were included if they had PCR-confirmed infection of SARS-CoV-2 and were not participating in a clinical evaluation of any Covid-19 pharmacotherapy.

We collected data on comorbidities and information about disease progression on daily resolution including breathing support required, oxygen flow, experimental anti-Covid-19 and antimicrobial drugs taken and several laboratory markers (PCR positivity for SARS-CoV-2, C-reactive protein, D-dimer, Interleukin 6, Ferritin, lymphocyte count). Full protocol for data collection is attached in the supplementary material. Due to very low number of patients using extra-corporeal membrane oxygenation (N = `r n_ecmo`) or non-invasive positive pressure ventilation (N = `r n_nippv`) in our sample, we collapsed those categories with mechanical ventilation.

Not all patients developed pneumonia or other symptoms of Covid-19. All patients received the standard of care including supplemental oxygen and ventilation, antibiotics for bacterial superinfections. Some patients were not indicated for all treatment modalities (especially mechanical ventilation) based on decision of attending physician and underlying patient condition. We note that the participating sites were not homogeneous in either patient population or treatment protocols. The choice of treatment was based on the decision of the attending clinician and availability.

@@

The character of the convenience sample does not allow for a proper assessment of
the association between treatments and patient outcomes,
because the treatments had not been assigned to patients at random, 
but were only observed retrospectively. This can be partially remedied by adjusting for patient characteristic in the analysis, but such adjustments will always be imperfect and the analysis  needs to be treated as exploratory and interpreted cautiously.

@@

First class of models are frequentist multistate models under the proportional hazards assumption as implemented in the `coxph` function from the `survival` package [@http://zotero.org/users/5567156/items/I9UVTUJF]. We primarily use a model with competing risks for death and discharge from hospital (see Figure&nbsp;\@ref(fig:modelstates)a).

Second class of models are Bayesian hidden Markov models of disease progression implemented via a custom extension to the `brms` package [@http://zotero.org/users/5567156/items/7ER3ZSER]. The parametrization of the HMM is inspired by [@http://zotero.org/users/5567156/items/F9K2L5Q8]: the actual process of disease is assumed to be continuous and allow only for transitions between neighboring states (as shown in Figure&nbsp;\@ref(fig:modelstates)b, c). The total probability of transition between any two states over the period of a day is then computed as the total probability of transition across all possible paths. This class of models does not satisfy the proportional hazards assumption, instead, it is assumed the process has the Markov property - i.e. that the (potentially unobserved) state and the covariates at a given day contain all the information to determine probabilities of the states on the next day. We use two versions of such models, one working solely with the observed breathing support and one assuming a hidden improving/worsening distinction. All of the hidden Markov models take into account whether best supportive care was initiated and a patient was thus not indicated to progress to more intensive treatment modalities.

@@

Except for age, sex and comorbidities, all covariates are treated as time-varying, e.g. the effect of taking a drug is only included for the days after the drug was taken.

@@

We searched the living systematic review of Covid-19 prognostic models [@http://zotero.org/users/5567156/items/TS57HJ3T] for models that could be applied to our dataset (i.e. where we have gathered all the input features).  We primarily focused on the Area Under Receiver Operating Characteristic Curve (AUC), and its bootstrapped 95% confidence intervals which we computed using the `pROC` package [@http://zotero.org/users/5567156/items/UN6BIC67]. When there were multiple reasonable ways to evaluate the outcome or a predictor in our dataset, we computed and reported all of those options. We used two simple scores with age or the decade of age as the sole predictor to have a simple baseline to compare the scores against.

@@

data_table_1 %>% group_by(group) %>% 
  summarise(
    N = n(),
    `Distinct sites` = length(unique(hospital_id)),
    Male = sum_percent(sex == "M"),
    `Age (mean, IQR)` = mean_iqr(age),
    `Admitted for Covid` = sum_percent(admitted_for_covid),
    `Took Azithromycin` = sum_percent(ever_az),
    `Took Dexamethasone` = sum_percent(ever_dexamethasone),
    `Took Favipiravir` = sum_percent(ever_favipiravir),
    `Took Remdesivir` = sum_percent(ever_remdesivir),
    `Convalescent plasma` = sum_percent(ever_convalescent_plasma),
    `Ischemic Heart Disease` = sum_percent(ischemic_heart_disease),
    `Takes antihypertensives` = sum_percent(has_hypertension_drugs),	
    `Heart Failure` = sum_percent(heart_failure),
    COPD = sum_percent(COPD),
    Asthma = sum_percent(asthma),
    `Other lung disease` = sum_percent(other_lung_disease),
    Diabetes = sum_percent(diabetes),
    `Renal Disease` = sum_percent(renal_disease),
    `Liver Disease` = sum_percent(liver_disease),
    `Smoking` = sum_percent(smoking),
    `BMI (mean, IQR)` = mean_iqr(BMI),
    `Best supportive care` = sum_percent(!is.na(best_supportive_care_from)),
    Deceased = sum_percent(outcome == "Death"),
    Discharged = sum_percent(outcome == "Discharged"),
    .groups = "drop"
  ) %>%
  mutate(group = factor(group, levels= c("All", "HCQ", "No HCQ", "Favipiravir"), ordered = TRUE)) %>%
  arrange(group) %>%
  as.data.frame() %>%
  column_to_rownames("group") %>%
  t() %>% 
  knitr::kable(caption = "Patient characteristics for the overall sample and treatment subgroups. Note that the Favipiravir subgroup is not exclusive with either the HCQ or No HCQ group. IQR = interquartile range, COPD = Chronic obstructive pulmonary disease, BMI = body-mass index, Best supportive care = patient was not indicated for progression to a more intensive treatment modality.", booktabs = TRUE)

@@

In Figure&nbsp;\@ref(fig:progression) we show the overall disease progression for all patients and in Figure&nbsp;\@ref(fig:selectedmarkers) we show the time-course of a subset of the markers we have measured. The data show some interesting patterns: patients with low Interleukin-6 or D-dimer values are overrepresented among patients with better outcomes, most patients were hospitalized with high CRP and for many the CRP levels stayed elevated over the whole hospitalization. The limited nature of the data however does not allow for any statistically robust conclusions. We also see that the marker levels were not substantially stratified by study site.  Those patterns should however be interpreted with care due to systematic missingness of the data - in particular, patients that fared worse were probably more likely to have the markers measured. However we believe this kind of patient-level view is useful to appreciate both the extent of between- and within-patient variability.

@@

As noted above, the nature of the convenience sample did not enforce random assignment of treatments to patients. To illustrate this phenomenon, we give an example of the association of treatment with Hydroxychloroquine on patients' baseline characteristics.

```{r hcqassoc, message=FALSE, fig.width=7, fig.height=4.5, fig.cap="The choice of treatment with Hydroxychloroquine seemed to be associated with the status of patients upon hospitalization. Comorbidities were associated with both worse outcome (black) and lower chance of treatment with Hydroxychloroquine (red). HCQ: hydroxychloroquine, IHD: ischemic heart disease, HD: hypertension drugs, HF: heart failure history, COPD: chronic obstructive pulmonary disease, LungD: other lung disease, Dia: diabetes, RenalD: renal disease, LiverD: liver disease, HighCr: creatinin above 115 for males or above 97 for females, HighInr: Prothrombin time (Quick test) as International Normalized Ratio above 1.2, LowAlb: albumin in serum/plasma below 36 g/l."}

tmp<-data$patient_data[,c('age','sex','days_from_symptom_onset','admitted_for_covid','ischemic_heart_disease','has_hypertension_drugs','heart_failure',
  'COPD','asthma','other_lung_disease','diabetes','renal_disease','liver_disease','smoking','BMI','obesity','last_record','high_creatinin','high_pt_inr','low_albumin',
  'heart_problems','outcome','ever_hcq','ever_az','worst_breathing','had_invasive'),]
tmp$outcome[tmp$outcome=='Transferred']<-'Hospitalized'
tmp$outcome<-dropLevels(tmp$outcome)
tmp2<-tmp[tmp$outcome%in%c('Death','Discharged'),]
tmp2$outcome<-dropLevels(tmp2$outcome)
tmp2$outcome<-tmp2$outcome=='Death'

m.ihd<-glm(outcome~ischemic_heart_disease,tmp2,family='binomial')
m.hd<-glm(outcome~has_hypertension_drugs,tmp2,family='binomial')
m.hf<-glm(outcome~heart_failure,tmp2,family='binomial')
m.copd<-glm(outcome~COPD,tmp2,family='binomial')
m.asthma<-glm(outcome~asthma,tmp2,family='binomial')
m.old<-glm(outcome~other_lung_disease,tmp2,family='binomial')
m.dia<-glm(outcome~diabetes,tmp2,family='binomial')
m.rd<-glm(outcome~renal_disease,tmp2,family='binomial')
m.ld<-glm(outcome~liver_disease,tmp2,family='binomial')
m.s<-glm(outcome~smoking,tmp2,family='binomial')
m.ob<-glm(outcome~obesity,tmp2,family='binomial')
m.hc<-glm(outcome~high_creatinin,tmp2,family='binomial')
m.hi<-glm(outcome~high_pt_inr,tmp2,family='binomial')
m.la<-glm(outcome~low_albumin,tmp2,family='binomial')

m2.ihd<-glm(ever_hcq~ischemic_heart_disease,tmp2,family='binomial')
m2.hd<-glm(ever_hcq~has_hypertension_drugs,tmp2,family='binomial')
m2.hf<-glm(ever_hcq~heart_failure,tmp2,family='binomial')
m2.copd<-glm(ever_hcq~COPD,tmp2,family='binomial')
m2.asthma<-glm(ever_hcq~asthma,tmp2,family='binomial')
m2.old<-glm(ever_hcq~other_lung_disease,tmp2,family='binomial')
m2.dia<-glm(ever_hcq~diabetes,tmp2,family='binomial')
m2.rd<-glm(ever_hcq~renal_disease,tmp2,family='binomial')
m2.ld<-glm(ever_hcq~liver_disease,tmp2,family='binomial')
m2.s<-glm(ever_hcq~smoking,tmp2,family='binomial')
m2.ob<-glm(outcome~obesity,tmp2,family='binomial')
m2.hc<-glm(ever_hcq~high_creatinin,tmp2,family='binomial')
m2.hi<-glm(ever_hcq~high_pt_inr,tmp2,family='binomial')
m2.la<-glm(ever_hcq~low_albumin,tmp2,family='binomial')

models<-list(m.ihd,m.hd,m.hf,m.copd,m.asthma,m.old,m.dia,m.rd,m.ld,m.s,m.ob,m.hc,m.hi,m.la)
modelNames<-c('IHD','HD','HF','COPD','Asthma','LungD','Dia','RenalD','LiverD','Smoking','Obesity','HighCr','HighInr','LowAlb')
ys<-c(0)
for (m in models) {
  ys<-c(min(min(ys),confint(m)[2,1]),max(max(ys),confint(m)[2,2]))
}
models2<-list(m2.ihd,m2.hd,m2.hf,m2.copd,m2.asthma,m2.old,m2.dia,m2.rd,m2.ld,m2.s,m2.ob,m2.hc,m2.hi,m2.la)
modelNames2<-modelNames
ys2<-c(0)
for (m in models2) {
  ys2<-c(min(min(ys2),confint(m)[2,1]),max(max(ys2),confint(m)[2,2]))
}

plot(ys,c(1,length(modelNames)),ty='n',yaxt='n',xlab='Log odds ratio',ylab='')
abline(v=0,col='blue')
col.hcq<-bfc()["Vermillion"]
for (i in seq(along=models)) {
  ii<-length(models)-i+1
  m<-models[[i]]
  m2<-models2[[i]]
  lines(confint(m)[2,],rep(ii,2))
  points(coef(m)[2],ii,pch=19)
  lines(-confint(m2)[2,],rep(ii,2)-.3,col=col.hcq)
  points(-coef(m2)[2],ii-.3,col=col.hcq,pch=19)
}
axis(2,length(modelNames):1,modelNames,las=2)
legend('right',c('Death','No HCQ'),lty=1,col=c('black',col.hcq))
```

Figure&nbsp;\@ref(fig:hcqassoc) shows that patients with worse baseline characteristics, which lead to worse outcome, 
were less likely to receive Hydroxychloroquine treatment. 
This clearly creates a bias towards a positive effect of Hydroxychloroquine on the outcome, 
which, however, could be false.

@@

Taken quantitatively, the comorbidities known upon hospitalization were informative with respect to future Hydroxychloroquine treatment:
the score representing the cumulative presence of ischemic heart disease, hypertension drugs, former heart failure, COPD, 
other lung diseases, renal disease, or high creatinin 
was associated with a lower chance of taking Hydroxychloroquine over the course of the hospitalization 
(the chance was only `r round(100*exp(coef(m.hcq_treatment2)[2]),1)`%, 95% confidence interval 
(`r round(100*exp(confint(m.hcq_treatment2)[2,]),1)`)%,
Chi-square test in the logistic regression model, 
$\chi^2$=`r round(abs(anova(m.hcq_treatment2,m.hcq_treatment0,test='Chisq'))[2,4],2)`, df=1, 
P=`r round(anova(m.hcq_treatment2,m.hcq_treatment0,test='Chisq')[2,5],3)`).

@@

```{r}
bayesian_cols <- cols(
  hypothesis = col_character(),
  model = col_character(),
  adjusted = col_character(),
  estimand = col_character(),
  widest_CI_excl_reference = col_double(),
  point_estimate = col_double(),
  ci_low = col_double(),
  ci_high = col_double()
)
res_bayesian <- 
  rbind(read_csv(here::here("manuscript", "hmm_res.csv"), col_types = bayesian_cols),
        read_csv(here::here("manuscript", "analysis_brms.csv"), col_types = bayesian_cols),
        read_csv(here::here("manuscript", "jm_res.csv"), col_types = bayesian_cols)
        )

frequentist_cols <- cols(
  hypothesis = col_character(),
  model = col_character(),
  adjusted = col_character(),
  p_value = col_double(),
  ci_low = col_double(),
  ci_high = col_double()
)

res_frequentist <- 
  rbind(read_csv(here::here("manuscript", "competing_risks_res.csv"), col_types = frequentist_cols),
        read_csv(here::here("manuscript", "single_outcome_coxph_res.csv"), col_types = frequentist_cols))

all_hypo <- rbind(res_bayesian %>% select(-widest_CI_excl_reference), 
      res_frequentist %>% select(-p_value)) %>%
    inner_join(hypotheses_df, by = c("hypothesis" = "name"))

all_hypo_drugs <- all_hypo  %>%
  filter(model_check == "OK")

@@

Here, we focus on Hydroxychloroquine and Azithromycin as those are the only treatment with larger sample size. We also investigate Favipiravir as it is less well reported in the literature. Hydroxychloroquine was dosed almost exclusively in a 5-day regime starting with a loading dose of 800mg on the first day and followed by 400mg. Majority of patients complemented Hydroxychloroquine with Azithromycin while Azithromycin was rarely used alone (see Table \@ref(tab:summarytable)). Azithromycin was most frequently dosed 250 or 500mg/day, but doses ranging from 100mg/day to 1500 mg/day were observed.  Favipiravir was used only at one site with a loading dose of 3600mg on the first day, followed by at most 9 days with a 1600mg dose. All but one of the patients receiving Favipiravir also received Hydroxychloroquine. Treatment was initiated mostly within two days of admission (see Figure&nbsp;\@ref(fig:initiation)).

```{r initiation, fig.cap="Histogram of timing of first treatment relative to admission into one of the study sites. Two patients initiated treatment before admission, which is shown as the negative numbers.", fig.height=2}
#data$marker_data %>% filter(marker == "favipiravir") %>% group_by(patient_id) %>% summarise(days = n(), .groups = "drop") %>% summarise(min_days = min(days), max_days = max(days))
#data$marker_data %>% filter(marker == "az") %>% group_by(patient_id) %>% summarise(days = sum(!is.na(value) & value > 0), .groups = "drop") %>% filter(days > 0) %>% ggplot(aes(x = days)) + geom_histogram(binwidth = 1)
#data$marker_data%>% filter(marker == "az") %>% group_by(value) %>% summarise(n())

#Patients treated with convalescent plasma received 1-3 units of plasma over 1-2 days.

>>>>

---
title: "Detailed disease progression of 213 patients hospitalized with Covid-19 in the Czech Republic: An exploratory analysis" 
output: 
  #rticles::arxiv_article
  #pdf_document
  # bookdown::pdf_document2:
  #    toc: false
  bookdown::word_document2: 
    toc: false
    number_sections: false
    reference_docx: "output_template.docx"
abstract: |  
  **Abstract:** We collected a multi-centric retrospective dataset of patients who were admitted to 10 hospitals in Czech Republic and tested positive for SARS-CoV-2. The dataset contains baseline patient characteristics, breathing support required, pharmacological treatment received and multiple markers on daily resolution. Patients in the dataset were treated with hydroxychloroquine (N = 108), azithromycin (N = 72), favipiravir (N = 9), convalescent plasma (N = 7), dexamethasone (N = 4) and remdesivir (N = 3), often in combination. Most patients were admitted during the first wave of the epidemic. To explore association between treatments and patient outcomes we performed multiverse analysis, observing how the conclusions change between defensible choices of statistical model, predictors included in the model and other analytical degrees of freedom.  Weak evidence to constrain the potential efficacy of azithromycin and favipiravir can be extracted from the data. Additionally we performed external validation of several proposed prognostic models for Covid-19 severity showing that they mostly perform unsatisfactorily on our dataset.
bibliography: Covid19.json
csl: journal-of-medical-virology.csl
date: "Version: `r format(Sys.time(), '%Y-%m-%d')`"
author: "`r { source(here::here('manuscript', 'authors.R')); get_authors_string() }`"
---

@@

The Covid-19 pandemic, caused by severe acute respiratory syndrome coronavirus (SARS-CoV-2) has, as of November 2020, led to over 46 million cases and over 1.1 million deaths. Proposed treatments include antivirals approved for other indications (chloroquine, hydroxychloroquine, lopinavir/ritonavir, remdesivir, favipiravir, umifenovir), azithromycin, corticosteroids, immunoglobulins, tocilizumab and convalescent plasma [@http://zotero.org/users/5567156/items/8XWTUMUI; @http://zotero.org/users/5567156/items/6Q8Z8VJR]. Currently, the only treatment that exhibited positive effects in randomized trials is the corticosteroid dexamethasone for severe cases [@http://zotero.org/users/5567156/items/IW6VZ53Q; @http://zotero.org/users/5567156/items/5VAKEW58]. Remdesivir, previously designed and approved for Ebola, SARS and MERS, is currently the only antiviral against SARS-CoV-2 approved by the FDA. However, during spring and summer it was available only in limited amounts via compassionate use programme. Randomized clinical trials on remdesivir have provided  conflicting evidence [@http://zotero.org/users/5567156/items/MFS3IPPU; @http://zotero.org/users/5567156/items/KB5UDZH3]. Despite initial promise, hydroxychloroquine has failed to show benefit for both hospitalized patients [@http://zotero.org/users/5567156/items/9X3UH8ZM; @http://zotero.org/users/5567156/items/KB5UDZH3], as early treatment [@http://zotero.org/users/5567156/items/G6KZKVMI], and as both pre- and post-exposure prophylaxis [@http://zotero.org/users/5567156/items/7C23YCV5; @http://zotero.org/users/5567156/items/GMR9FX9Z; @http://zotero.org/users/5567156/items/H4CLJNUR]. Further, no improvement could be discerned when adding azithromycin to hydroxychloroquine therapy [@http://zotero.org/users/5567156/items/BQLRZGTQ]. Similarly, Lopinavir/Ritonavir also did not show benefit [@http://zotero.org/users/5567156/items/6I6GAVWU]. No benefit was observed for convalescent plasma [@http://zotero.org/users/5567156/items/UEXLDJCB; @http://zotero.org/users/5567156/items/BHGT88US].

Interestingly, in observational studies, hydroxychloroquine was often found to be associated with better outcomes [e. g., @http://zotero.org/users/5567156/items/WZPNBRYN; @http://zotero.org/users/5567156/items/4HR7MUJ6; @http://zotero.org/users/5567156/items/NXC9CDI9].

Favipiravir is one of the less explored experimental treatments, previously used for influenza including avian types, with only one larger RCT in Covid- 19 known to us [@http://zotero.org/users/5567156/items/ISLSL7UC], the study however has many potentially severe problems [@http://zotero.org/users/5567156/items/7HDXLWEA]. Even considering studies or case series with only several treated patients, very few are available [@http://zotero.org/users/5567156/items/F3VNTQ92; @http://zotero.org/users/5567156/items/FSUQMUP8; @http://zotero.org/users/5567156/items/894H9N9B]. None of the results allow strong conclusions about benefits or harms of favipiravir, although a  metaanalysis of all available small studies showed an effect of favipiravir on clinical improvement in mild and moderate disease [@http://zotero.org/users/5567156/items/7595KBWR].

 For additional summary of treatments under investigation see [@http://zotero.org/users/5567156/items/YP62P6XJ].

@@

# Patients and Methods

@@

n_ecmo <- data$marker_data_wide %>% filter(breathing == "ECMO") %>% pull(patient_id) %>% unique() %>% length()
n_nippv <- data$marker_data_wide %>% filter(breathing == "NIPPV") %>% pull(patient_id) %>% unique() %>% length()

n_second_wave <- data$patient_data %>% filter(!first_wave) %>% nrow()
```

A convenience sample of patients from `r n_hospitals` sites was collected. For each site, the dataset contains all patients hospitalized in the participating wards over the data collection period. The data collection started at the onset of the Covid-19 pandemic (except for one site where some older records were inaccessible), but the end date for collection differs between sites due to time constraints of the participating physicians. Three sites included total of `r n_second_wave` patients that could be considered part of "second wave", i.e. admitted after September 1st, last patient included in the dataset was admitted on October 12th. See Supplementary Figure 1 for per-site data collection periods. Patients over the age of 18 were included if they had PCR-confirmed infection of SARS-CoV-2 and were not participating in a clinical trial of any Covid-19 pharmacotherapy.

We collected data on comorbidities and information about disease progression on daily resolution including breathing support required, oxygen flow, experimental anti-Covid-19 and antimicrobial drugs taken and several laboratory markers (PCR positivity for SARS-CoV-2, C-reactive protein, D-dimer, Interleukin 6, Ferritin, lymphocyte count). Full protocol for data collection is attached in the supplementary material. Due to very low number of patients using extra-corporeal membrane oxygenation (N = `r n_ecmo`) or non-invasive positive pressure ventilation (N = `r n_nippv`) in our sample, we merged those categories with mechanical ventilation.

Not all patients developed pneumonia or other symptoms of Covid-19. All patients received the standard of care including supplemental oxygen and ventilation, antibiotics for bacterial superinfections. Some patients were not indicated for all treatment modalities (especially mechanical ventilation) based on decision of the attending physician and underlying patient condition. We note that the participating sites were not homogeneous in either patient population or treatment protocols. The choice of treatment was based on the decision of the attending clinician and availability.

@@

The character of the convenience sample does not allow for a proper assessment of
the association between treatments and patient outcomes,
because the treatments had not been assigned to patients at random, 
but were only observed retrospectively. This can be partially remedied by adjusting for patient characteristics in the analysis, but such adjustments will always be imperfect and the analysis  needs to be treated as exploratory and interpreted cautiously.

@@

First class of models are frequentist survival and multistate models under the proportional hazards assumption as implemented in the `coxph` function from the `survival` package [@http://zotero.org/users/5567156/items/I9UVTUJF]. We primarily use a model with competing risks for death and discharge from hospital (see Figure&nbsp;\@ref(fig:modelstates)a).

Second class of models are Bayesian hidden Markov models (HMM) of disease progression implemented via a custom extension to the `brms` package [@http://zotero.org/users/5567156/items/7ER3ZSER]. The parametrization of the HMM is inspired by Williams et al.[@http://zotero.org/users/5567156/items/F9K2L5Q8]: the actual process of disease is assumed to be continuous and allow only for transitions between neighboring states (as shown in Figure&nbsp;\@ref(fig:modelstates)b, c). The total probability of transition between any two states over the period of a day is then computed as the total probability of transition across all possible paths. This class of models does not satisfy the proportional hazards assumption, instead, it is assumed the process has the Markov property - i.e. that the (potentially unobserved) state and the covariates at a given day contain all the information to determine probabilities of the states on the next day. We use two versions of such models, one working solely with the observed breathing support and one assuming a hidden improving/worsening distinction. All of the hidden Markov models take into account whether best supportive care was initiated and a patient was thus not indicated to progress to more intensive treatment modalities.

@@

Except for age, sex and comorbidities, all covariates are treated as time-varying, e.g., the effect of taking a drug is only included for the days after the drug was taken. More details on the exact model formulations can be found in the supplementary statistical analysis.

@@

We searched the living systematic review of Covid-19 prognostic models [@http://zotero.org/users/5567156/items/TS57HJ3T] for those that could be applied to our dataset (i.e. where we have gathered all the input features).  We primarily focused on the Area Under Receiver Operating Characteristic Curve (AUC), and its bootstrapped 95% confidence intervals which we computed using the `pROC` package [@http://zotero.org/users/5567156/items/UN6BIC67]. When there were multiple reasonable ways to evaluate the outcome or a predictor in our dataset, we computed and reported all of those options. We used two simple scores with age or the decade of age as the sole predictor to have a baseline to compare the scores against.

@@

data_table_1 %>% group_by(group) %>% 
  summarise(
    N = n(),
    `Distinct sites` = length(unique(hospital_id)),
    Male = sum_percent(sex == "M"),
    `Age (mean, IQR)` = mean_iqr(age),
    `Admitted for Covid` = sum_percent(admitted_for_covid),
    `Took azithromycin` = sum_percent(ever_az),
    `Took dexamethasone` = sum_percent(ever_dexamethasone),
    `Took favipiravir` = sum_percent(ever_favipiravir),
    `Took remdesivir` = sum_percent(ever_remdesivir),
    `Convalescent plasma` = sum_percent(ever_convalescent_plasma),
    `Ischemic Heart Disease` = sum_percent(ischemic_heart_disease),
    `Takes antihypertensives` = sum_percent(has_hypertension_drugs),	
    `Heart Failure` = sum_percent(heart_failure),
    COPD = sum_percent(COPD),
    Asthma = sum_percent(asthma),
    `Other lung disease` = sum_percent(other_lung_disease),
    Diabetes = sum_percent(diabetes),
    `Renal Disease` = sum_percent(renal_disease),
    `Liver Disease` = sum_percent(liver_disease),
    `Smoking` = sum_percent(smoking),
    `BMI (mean, IQR)` = mean_iqr(BMI),
    `Best supportive care` = sum_percent(!is.na(best_supportive_care_from)),
    Deceased = sum_percent(outcome == "Death"),
    Discharged = sum_percent(outcome == "Discharged"),
    .groups = "drop"
  ) %>%
  mutate(group = factor(group, levels= c("All", "HCQ", "No HCQ", "Favipiravir"), ordered = TRUE)) %>%
  arrange(group) %>%
  as.data.frame() %>%
  column_to_rownames("group") %>%
  t() %>% 
  knitr::kable(caption = "Patient characteristics for the overall sample and treatment subgroups. Note that the favipiravir subgroup is not exclusive with either the HCQ or No HCQ group. IQR = interquartile range, COPD = Chronic obstructive pulmonary disease, BMI = body-mass index, Best supportive care = patient was not indicated for progression to a more intensive treatment modality.", booktabs = TRUE)

@@

In Figure&nbsp;\@ref(fig:progression) we show the overall disease progression for all patients and in Figure&nbsp;\@ref(fig:selectedmarkers) we show the time-course of a subset of the markers we have measured. The data show some interesting patterns: patients with low Interleukin-6 or D-dimer values are overrepresented among patients with better outcomes, most patients had high CRP upon admission and for many the CRP levels stayed elevated over the whole hospitalization. The limited nature of the data however does not allow for any statistically robust conclusions. We also see that the marker levels were not substantially stratified by study site.  Those patterns should however be interpreted with care due to systematic missingness of the data - in particular, patients that fared worse were probably more likely to have the markers measured. However we believe this kind of patient-level view is useful to appreciate the extent of both between-patient and within-patient variability.

@@

As noted above, the nature of the convenience sample did not enforce random assignment of treatments to patients. In fact, patients with worse baseline characteristics, which lead to worse outcomes, were less likely to receive hydroxychloroquine (see Supplementary Figure 3). This clearly creates a bias towards a positive effect of hydroxychloroquine on the outcome (and potentially for other treatments as well - most were used in combination with hydroxychloroquine), which, however, could be false.

@@

Taken quantitatively, the comorbidities known upon hospitalization were informative with respect to future hydroxychloroquine treatment:
the score representing the cumulative presence of ischemic heart disease, hypertension drugs, former heart failure, COPD, 
other lung diseases, renal disease, or high creatinine 
was associated with a lower chance of taking hydroxychloroquine over the course of the hospitalization 
(the chance was only `r round(100*exp(coef(m.hcq_treatment2)[2]),1)`%, 95% confidence interval 
(`r round(100*exp(confint(m.hcq_treatment2)[2,]),1)`)%,
Chi-square test in the logistic regression model, 
$\chi^2$=`r round(abs(anova(m.hcq_treatment2,m.hcq_treatment0,test='Chisq'))[2,4],2)`, df=1, 
P=`r round(anova(m.hcq_treatment2,m.hcq_treatment0,test='Chisq')[2,5],3)`).

@@

```{r}

all_hypo_drugs <- read_all_multiverse_res()  %>%
  filter(model_check == "OK", !adj.first_wave, estimand_group == "")

first_fig_multiverse_drugs <- 4
last_fig_multiverse_drugs <- 6
supplementary_figure_markers <- last_fig_multiverse_drugs + 1 
supplementary_figures_multiverse <-  paste0(first_fig_multiverse_drugs, "-", last_fig_multiverse_drugs)

@@

Here, we focus on hydroxychloroquine and azithromycin as those are the only treatments with larger sample size. We also investigate favipiravir as it is less well reported in the literature. Hydroxychloroquine was dosed almost exclusively in a 5-day regime starting with a loading dose of 800mg on the first day and followed by 400mg. Majority of patients complemented hydroxychloroquine with azithromycin while azithromycin was rarely used alone (see Table \@ref(tab:summarytable)). Azithromycin was most frequently dosed 250 or 500mg/day, but doses ranging from 100mg/day to 1500 mg/day were observed.  Favipiravir was used only at one site with a loading dose of 3600mg on the first day, followed by at most 9 days with a 1600mg dose. All but one of the patients receiving favipiravir also received hydroxychloroquine. Treatment was initiated mostly within two days of admission (see Supplementary Figure 2).

The results of the multiverse analysis for association between hydroxochloroquine, azithromycin and favipiravir usage and death is shown in Figure&nbsp;\@ref(fig:multiverseresultsdrugs) - here, we only show models that were not found to have immediate problems representing the data well or computational issues. Results for all models we tested are reported in  supplementary Figures `r supplementary_figures_multiverse`, with additional details in supplementary statistical analysis. The results do not change noticeably when only patients from the first wave are included (supplementary Figures `r supplementary_figures_multiverse`).
